---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## **Introduction**

There is a well-known relationship between the consumption of fruits and vegetables and multiple health benefits (lower blood pressure, reduce the risk of heart disease and stroke, prevention of some types of cancer, etc.). The main objective of this assessment is to analyze to what extent the consumption of fruit a vegetables can be associated with a good level of health in adults. Also, it will be analyzed if this relationship between consumption and health is preserved when we control some confounding variables. 

To assess these objectives it will be used the data of The Behavioral Risk Factor Surveillance System (BRFSS). The complete information about this dataset can be found [here](https://d3c33hcgiwev3.cloudfront.net/_e34476fda339107329fc316d1f98e042_brfss_codebook.html?Expires=1615593600&Signature=V4ZmP7Ff5hinmnyxLsCv8j0F3tXR5hHjIlb0oBS64~4ZqhQ4o3D0S9n-uJ0fQ6VAVaZMgnvtTiRg5Yrg5TQorIoiGKBCyCAyKHzlrwTf0i6oYRJ~wCvTUm-HBJ2-P8qcJr7u3fQbZXLgP1Z-uli2GTopBAd-2sMJ0j~7UiJAwbU_&Key-Pair-Id=APKAJLTNE6QMUY6HBC5A). The limitations of this dataset will be explained in the data part.

## **Setup**

### **Load packages**

Four packages were used to conduct the data analysis. If you want to reproduce it, please make sure that they are previously installed.

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(kableExtra)

```

### Load data

```{r load-data}
load("brfss2013.RData")
```

* * *

## **Part 1: Data**

The data of The Behavioral Risk Factor Surveillance System (BRFSS) is collected by a Stratified sample method, where the strata are the states. For instance, there is a random sample within each state. In this sense, the conclusions derived from the use of the dataset are  ***generalizable*** to the non-institutionalized adult population of the US. However, the survey doesn´t contemplate a random assignment of treatment. 

So we can conclude that the BRFSS is useful only to conduct ***observational studies***, where we can only assess association,  ***not causality***.

Despite the generalizability of the survey, the method of collecting the data could imply some biases. Cause the collection relies on a landline telephone and cellular telephone-based surveys, the results of the survey exclude the adults that don´t have access to a landline or cellular telephone.

Also, there are some disadvantages of the self-report approach of the BRFSS. Individuals being asked a sensitive and complex question may have limitations in answering with neutrality and objectivity.

* * *

## **Part 2: Research questions**

**Research question 1:** 
Is there a relationship between the **health quality** and the level of 
**consumption of fruit and vegetables**?

Using the concepts of the course, we can rewrite this question as: *the probability of having good or poor health depends on the level consumption of fruit and vegetables?*

**Research question 2:** 
Can we find possible confounding variables that affect the relationship between quality of health and level of consumption of fruit and vegetables? are **income** and **physical activity** candidates? 

One of the reasons that prevent us assess causality in the first question is that the BRFSS doesn´t control confounders. We want to prove that the variables of income and physical activity are indeed affecting the relationship.

**Research question 3:**
If we control confounding variables, the relationship between **health quality** and **consumption of fruit and vegetables** still prevails?

This question does not pretend to find causality. The purpose is to see if the relationship is still present when we take into account the confounding variables.

* * *

## **Part 3: Exploratory data analysis**

### **Research question 1:**

To assess the relationship between **health quality** and **level of consumption of fruit and vegetables** is needed some transformation of the data. In the code chunk below, we create a continuous variable that shows the total consumption of fruit and vegetables for each of the respondents. Then, we recategorized the variable to create a discrete one. 

```{r,  message=FALSE}
levels(brfss2013$X_incomg) <- c("< $15,000", 
                                    "$15,000 to $25,000",
                                    "$25,000 to $35,000",
                                    "$35,000 to $50,000",
                                    "$50,000 > ")

levels(brfss2013$exerany2) <- c("Physical activity in past 30 day", 
                                "No physical activity in past 30 day")

brfss2013Food <- brfss2013 %>% 
        filter(fruitju1 %in% (301:399) &
                       fruit1 %in% (301:399) &
                       fvgreen %in% (301:399) &
                       fvorang %in% (301:399) &
                       vegetab1 %in% (301:399))  %>% 
        mutate(totFruitVegetables = 
                       fruitju1 + fruit1 + fvgreen + fvorang + vegetab1-1500) %>%
        mutate(averageDayFruitVeg = totFruitVegetables/30.4167)


brfss2013Food <- brfss2013Food %>%
        mutate(fruitVegeGroups = 
                       cut(brfss2013Food$averageDayFruitVeg,breaks = c(0,1,3,16.27395)))

levels(brfss2013Food$fruitVegeGroups) <- c("> 1 fruit/vegetable per day", 
                                       " 1 to 3 fruit/vegetable per day",
                                       "< 3 fruit/vegetable per day")

```

Using the continuous variable of consumption of fruit/vegetables we can calculate the mean and the standard deviation:

```{r,  message=FALSE}
mean(brfss2013Food$averageDayFruitVeg)
sd(brfss2013Food$averageDayFruitVeg)
```

Using the discrete variable of consumption of fruit/vegetables we can calculate the marginal probabilities of each level of consumption.

```{r,  message=FALSE}
t <- brfss2013Food  %>%  filter(!is.na(brfss2013Food$fruitVegeGroups))  %>%
                             group_by(fruitVegeGroups) %>%
                             summarise(count = n()) %>% 
                             mutate(marginal_probability = count/sum(count))
t %>% kbl() %>% kable_styling()

```
Also,  we can calculate the marginal probabilities of the variable of quality of health
```{r,  message=FALSE}
t <- brfss2013Food  %>%  filter(!is.na(brfss2013Food$X_rfhlth))  %>%
                             group_by(X_rfhlth) %>%
                             summarise(count = n()) %>% 
                             mutate(marginal_probability = count/sum(count))
t %>% kbl() %>% kable_styling()

```

A stacked barplot is a good way to analyze the relations between health quality and level of consumption (in it´s discrete form).

```{r ,  message=FALSE, fig.align='center', fig.width= 7, fig.height = 4}

plot1 <- brfss2013Food %>%
        filter(!is.na(brfss2013Food$X_rfhlth), !is.na(brfss2013Food$fruitVegeGroups)) %>%
        group_by(fruitVegeGroups,X_rfhlth) %>%
        summarize(n = n()) %>%
        mutate(freq = n/sum(n),
               posn_pct = cumsum(freq)-0.5*freq)
names(plot1)[2] <- c("Health_Status")


ggplot(plot1, aes(fill = Health_Status , y = freq , x = fruitVegeGroups)) + 
        geom_bar(position ="stack", stat = "identity", width = .8, color="black") +
        geom_text(aes(label= round(freq, digits = 2), y=1-posn_pct), color="white")+
        ggtitle("Probability of having good or poor healt given the daily 
             consumption of fruit and vegetable") +
        ylab("Relative frequency") + 
        scale_fill_manual(values=c("steelblue2", "indianred2")) + 
        theme(axis.text.x=element_text(size=8), axis.title.x=element_blank(), 
              plot.title = element_text(hjust = 0.5)) 


```

The plot shows the conditional probabilities of having good or poor health given the level of daily consumption of fruit and vegetable. Using the notation of the class, we can see that ...   
    
<center>
**P(Good or better health | more that 3 fruit/vegetable per day) > P(Good or better health| less than 1 fruit/vegetable per day)**    
</center>     

* * *
### **Research question 2:**

We have the hypothesis that **income** and **physical activity** are confounding variables to the relationship of the first question. 

In order to solve this question its necessary to prove that the variables are associated with both **health** and **consumption of fruit and vegetables**

Let´s start with the income variable. 

We can calculate the marginal probabilities of the of income variable
```{r,  message=FALSE}
t <- brfss2013Food  %>%  filter(!is.na(brfss2013Food$X_incomg))  %>%
                             group_by(X_incomg) %>%
                             summarise(count = n()) %>% 
                             mutate(marginal_probability = count/sum(count))
t %>% kbl() %>% kable_styling()

```

The code chunk below constructs two stacked barplots where the income variable interacts, first, with the level of consumption of fruit/vegetables, then, with the health variable.

```{r, fig.align='center', message=FALSE, fig.width= 7, fig.height = 8}
##########################################################################################
plot2 <- brfss2013Food %>%
        filter(!is.na(brfss2013Food$fruitVegeGroups), !is.na(brfss2013Food$X_incomg)) %>%
        group_by(X_incomg,fruitVegeGroups) %>%
        summarize(n = n()) %>%
        mutate(freq = n/sum(n),
               posn_pct = cumsum(freq)-0.5*freq)
names(plot2)[2] <- c("Level_of_consumption")


p2 <- ggplot(plot2, aes(fill = Level_of_consumption , y = freq , x = X_incomg)) + 
            geom_bar(position ="stack", stat = "identity",
                     width = .8, color="black") +
            geom_text(aes(label= round(freq, digits = 2),
                          y=1-posn_pct), color="white") +
            ggtitle("Probability of having a specific level of
            fruit/vegetable consumption
            given the level of income") +
            ylab("Relative frequency") + 
            theme(axis.text.x=element_text(size=8),
                  axis.title.x=element_blank(), 
                  plot.title = element_text(hjust = 0.5)) 
##########################################################################################
plot3 <- brfss2013 %>%
        filter(!is.na(brfss2013$X_rfhlth), !is.na(brfss2013$X_incomg)) %>%
        group_by(X_incomg,X_rfhlth) %>%
        summarize(n = n()) %>%
        mutate(freq = n/sum(n),
               posn_pct = cumsum(freq)-0.5*freq)
names(plot3)[2] <- c("Health_Status")


p3 <- ggplot(plot3, aes(fill = Health_Status , y = freq , x = X_incomg)) + 
                geom_bar(position ="stack", stat = "identity",
                         width = .8, color="black") +
                geom_text(aes(label= round(freq, digits = 2), 
                              y=1-posn_pct), color="white") +
                ggtitle("Probability of having good or poor healt given
                         the level of income") +
                ylab("Relative frequency") + 
                scale_fill_manual(values=c("steelblue2", "indianred2")) + 
                theme(axis.text.x=element_text(size=8),
                      axis.title.x=element_blank(),
                      plot.title = element_text(hjust = 0.5)) 
        
grid.arrange(p2, p3, ncol=1)

```

As we can see, the income variable is associated with the level con consumption of fruit/vegetables. Using the probability notation...
      
<center>
**P(eating more than 3 fruit/vegetables | income greater than $50.000) > P(eating more than 3 fruit/vegetables | income less than $15.000)**  
       
</center>     
    
also, the income variable is associated with the health variable...         
<center> 
     
**P(Good or better health | income greater than $50.000) > P(Good or better health | income less than $15.000)** 
       
</center>
     
So we can determine that income is a confounding variable.
  
* * *    
    
Now, let´s do the same analysis for the variable of physical activity. We will use a continuous (Number of physical activities realized in the last month) and discrete versions (Dummy variable)

```{r,  message=FALSE}
brfss2013FoodExce <- brfss2013Food %>% 
        filter(exeroft1 %in% (201:299)) %>% 
        mutate(ExerciPerMonth = 
                       exeroft1 - 200 )
```

Using the continuous variable of physical activity we can calculate the mean and the standard deviation:

```{r,  message=FALSE}

mean(brfss2013FoodExce$ExerciPerMonth)
sd(brfss2013FoodExce$ExerciPerMonth)
```

Using the discrete variable of physical activity  we can calculate the marginal probabilities of each level of consumption.

```{r,  message=FALSE}
t <- brfss2013 %>%  filter(!is.na(brfss2013$exerany2))  %>%
                             group_by(exerany2) %>%
                             summarise(count = n()) %>% 
                             mutate(marginal_probability = count/sum(count))
t %>% kbl() %>% kable_styling()

```

The code chunk below constructs a scatterplot and a stacked barplot. The scatterplot shows the interaction between the realization of physical activity (in the continuous version of the variable) and the consumption of fruit and vegetables ( in the continuous version of the variable). The barplot shows the interaction between the realization of physical activity (in the discrete version of the variable) and the health status.

```{r, fig.align='center', message=FALSE, fig.width= 7, fig.height = 8}
##########################################################################################
p4 <- ggplot(brfss2013FoodExce, aes(x=ExerciPerMonth, y=averageDayFruitVeg)) +
        geom_point(alpha = 1/5, size = 1.5) +
        geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
        labs(x = "Number of physical activities realized in the last month",
             y = "Daily consumption (units of fruit/vegetables)")


plot5 <- brfss2013 %>%
        filter(!is.na(brfss2013$X_rfhlth), !is.na(brfss2013$exerany2)) %>%
        group_by(exerany2,X_rfhlth) %>%
        summarize(n = n()) %>%
        mutate(freq = n/sum(n),
               posn_pct = cumsum(freq)-0.5*freq)
names(plot5)[2] <- c("Health_Status")

##########################################################################################
p5 <- ggplot(plot5, aes(fill = Health_Status , y = freq , x = exerany2)) + 
      geom_bar(position ="stack", stat = "identity", width = .8, color="black") +
      geom_text(aes(label= round(freq, digits = 2), y=1-posn_pct), color="white") +
      ggtitle("Probability of having good or poor healt given
                             the realization of exercise") +
      ylab("Relative frequency") + 
      scale_fill_manual(values=c("steelblue2", "indianred2")) + 
      theme(axis.text.x=element_text(size=8), axis.title.x=element_blank()) 

grid.arrange(p4, p5, ncol=1)

```

As we can see in the scatterplot, the physical activity variable has a positive correlation with the level of consumption of fruit/vegetables. In this sense, a person that practices a lot of physical activity also has a high consumption of fruit and vegetables.

Rewriting this idea with the probability notation (and assuming that the variables have discrete levels)...
   
<center>
**P(high consumption of fruit/vegetables | a lot of physical activity) > P(high consumption of fruit/vegetables | little physical activity)**  
       
</center>  
   
Also, for the stacked barplot we can see that the physical activity is associated with the health variable....
   
<center> 
     
**P(Good or better health | physical activity in past 30 days) > P(Good or better health | no physical activity in past 30 days)** 
   
</center>  

So we can determine that physical activity is a confounding variable.  
       
* * *
### **Research question 3:**

In this question, we would control the confounding variables and check if the association between health quality and consumption of fruit and vegetables still prevails.

Let´s start trying to control the income variable.

Again a stacked barplot is a good way to visualize the association *across* the income levels.   

```{r,  fig.align='center', message=FALSE, fig.width= 12, fig.height = 8}

##########################################################################################
plot1 <- brfss2013Food %>%
        filter(!is.na(brfss2013Food$X_rfhlth), 
               !is.na(brfss2013Food$fruitVegeGroups),
               !is.na(brfss2013Food$X_incomg)) %>%
        group_by(X_incomg,fruitVegeGroups,X_rfhlth) %>%
        summarize(n = n()) %>%
        mutate(freq = n/sum(n), posn_pct = cumsum(freq)-0.5*freq)


names(plot1)[3] <- c("Health_Status")

ggplot(plot1, aes(fill = Health_Status , y = freq , x = fruitVegeGroups)) + 
      geom_bar(position ="stack", stat = "identity", width = .8, color="black") +
      geom_text(aes(label= round(freq, digits = 2), y=1-posn_pct), color="white") +
      facet_grid( X_incomg~.)+
      ggtitle("Probability of having good or poor healt given the daily consumption of fruit and vegetable (across income levels)") +
      ylab("Relative frequency") + 
      scale_fill_manual(values=c("steelblue2", "indianred2")) + 
      theme(axis.text.x=element_text(size=10), 
            axis.title.x=element_blank(),
            plot.title = element_text(hjust = 0.5)) 




```
Across the income levels the relationship of the first question still prevails: a high level of consumption of fruit/vegetables is associated with better health. This is true for both individuals of high income and individuals of low income.
* * *

Now let´s control the physical activity variable.

In this case, the best option is to use the continuous version of the variable of consumption of fruit and vegetables and present a boxplot. 

To determine the effect of controlling the physical activity variable, we present a boxplot that does not consider this variable.

```{r,  fig.align='center', message=FALSE, fig.width= 15, fig.height = 12, message=FALSE, warning=FALSE}
##########################################################################################
plot2 <- brfss2013Food %>% 
  filter(!is.na(brfss2013Food$averageDayFruitVeg), 
         !is.na(brfss2013Food$X_rfhlth),
         averageDayFruitVeg < 5)
names(plot2)[which(names(plot2) == "X_rfhlth")] <- c("Health_Status")

M <- plot2 %>% group_by(Health_Status) %>% 
  summarise(means= mean(averageDayFruitVeg))

means <- M$means

p1 <- ggplot(plot2, aes(x=Health_Status, y=averageDayFruitVeg)) +
      geom_boxplot(aes(fill=Health_Status)) +
      stat_summary(fun=mean, geom="point", position=position_dodge(.9),show.legend=FALSE, 
                   color="black", size= 5) +
      scale_fill_manual(values=c("steelblue2", "indianred2")) +  
      geom_text(data = M, aes(y = means + 0.2, label=round(means,2)),
                position=position_dodge(1.2)) + 
      labs(x = "Health Satus",
           y = "Daily consumption (units of fruit/vegetables)",
           title = "Without controlling the physical activity variable")
    
##########################################################################################
plot2 <- brfss2013Food %>% 
  filter(!is.na(brfss2013Food$averageDayFruitVeg), 
         !is.na(brfss2013Food$X_rfhlth),
         !is.na(brfss2013Food$exerany2), averageDayFruitVeg < 5)
names(plot2)[which(names(plot2) == "X_rfhlth")] <- c("Health_Status")

M <- plot2 %>% group_by(exerany2,Health_Status) %>% 
  summarise(means= mean(averageDayFruitVeg))

means <- M$means

p2 <- ggplot(plot2, aes(x=exerany2, y=averageDayFruitVeg)) +
      geom_boxplot(aes(fill=Health_Status), position=position_dodge(.9)) +
      stat_summary(fun=mean, geom="point", aes(group=Health_Status), position=position_dodge(.9),show.legend=FALSE, 
                   color="black", size= 5) +
      scale_fill_manual(values=c("steelblue2", "indianred2")) +  
      geom_text(data = M, aes(y = means, label=round(means,2), group=Health_Status),
                 position=position_dodge(1.4)) + 
      labs(x = "physical activities in the last month",
           y = "Daily consumption (units of fruit/vegetables)",
           title = "Controlling the physical activity variable")


grid.arrange(p1, p2, ncol=2)


```

The first boxplot shows the same result of the first question with a different logic. Its usefulness lies in contrasting it with the boxplot that controls the physical activity variable.

When we control the confounding variable, the relationship between health and consumption can only be seen in the group that realized physical activity. The relationship is broken in the group that didn´t realize physical activity.

* * *
## **Conclusion**

In this assessment we explored the relationship between the level of consumption of fruit and vegetables and the health status using the dataset of the BRFSS. Using stacked barplots, we observed that there is a positive association between the two variables. However, we also determined that there are confounding variables that affect this association. We focus on the income and physical activity variables. Finally, when we tried to control these variables we found contrasting results: the relationship still prevails when we control the income variable, but it was partially broken when we control the physical activity. These results underline the idea that the association of two variables doesn't imply causality.

